---
title: "Cure model appropriateness"
output:
  html_document:
    self_contained: true
    theme: null
    highlight: null
    mathjax: null
    toc: true
    toc_depth: 2
params:
  label: NULL
  source: NULL
  map: NULL
  prepared: NULL
  fit: NULL
  assess: NULL
  include_lognormal: FALSE
  expert_confirmed: FALSE
---

<!--
FIXPASS: report/report.Rmd did not exist. The downloadable HTML report was in
scope (docs/shiny-app-spec.md section 7) and was a Definition-of-Done item, but
no file was ever created and mod_conclusion.R had no downloadHandler. This is
that file.

THE CARDINAL RULE (R1) HOLDS HERE TOO. This document computes no inferential
statistic. Every inferential number below is read out of a field of an object
returned by an exported cureAssess function, and every one of them carries a
provenance line naming the call and the field. The only arithmetic performed in
this file is the app's own permitted list: the six descriptive data counts
(n, events, censored %, median follow-up, max follow-up, last event time) and
the single closed form 1 - alpha^(1/n) for the qn threshold, which qn.test()
does not return as a field.

The copy constants in app/R/helpers.R are deliberately NOT sourced. The app
renders this Rmd from a copy in tempdir() with knit_root_dir set there, so no
relative path reaches app/R/helpers.R, and an absolute path would break the
promise that this file also knits by hand on any machine. The verbatim copy
blocks below are therefore inlined, and they are copied from the build contract
section J, the same source helpers.R was filled from.

The CSS is inlined for the same reason: the output must be one self-contained
file that can be emailed and opened offline, so it never links app/www/app.css.
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 9, fig.height = 6.5, dpi = 110
)

# ---- what did we actually get? ---------------------------------------------
# Every param has a NULL-safe default so this file knits by hand with no app
# running. Each section below is gated on one of these flags and degrades to a
# plain sentence rather than erroring.
P_PREPARED <- if (is.data.frame(params$prepared) && nrow(params$prepared) > 0L) params$prepared else NULL
P_ASSESS   <- params$assess
P_FIT      <- params$fit
P_MAP      <- params$map

HAVE_PREP   <- !is.null(P_PREPARED)
HAVE_ASSESS <- !is.null(P_ASSESS)
HAVE_TESTS  <- HAVE_ASSESS && isTRUE(P_ASSESS$tests_run) && !is.null(P_ASSESS$tests)
HAVE_AIC    <- HAVE_ASSESS && is.data.frame(P_ASSESS$screening$aic_table)
HAVE_KM     <- !is.null(P_FIT) && !is.null(P_FIT$kmplot)
HAVE_ANY    <- HAVE_PREP || HAVE_ASSESS

TESTS <- if (HAVE_TESTS) P_ASSESS$tests else NULL

# ---- formatting, identical in behaviour to helpers.R F.5-F.7 ---------------
rp_dash <- function() "—"

rp_num <- function(x, digits = 4) {
  if (is.null(x) || length(x) != 1L) return(rp_dash())
  v <- suppressWarnings(as.numeric(x))
  if (length(v) != 1L || !is.finite(v)) return(rp_dash())
  if (v != 0 && abs(v) < 1e-4) return(format(v, digits = 3, scientific = TRUE))
  formatC(v, format = "f", digits = digits)
}

rp_pct <- function(x, digits = 1) {
  if (is.null(x) || length(x) != 1L) return(rp_dash())
  v <- suppressWarnings(as.numeric(x))
  if (length(v) != 1L || !is.finite(v)) return(rp_dash())
  paste0(formatC(v, format = "f", digits = digits), "%")
}

# A length-1 finite numeric out of a package field, or NA. Package statistics
# come back NA when a diagnostic is not computable and NULL when tests_run is
# FALSE; both must land on an em dash, never on the text "NA".
rp_stat <- function(x) {
  if (is.null(x) || length(x) != 1L) return(NA_real_)
  v <- suppressWarnings(as.numeric(x))
  if (length(v) != 1L || !is.finite(v)) NA_real_ else v
}

esc <- function(x) htmltools::htmlEscape(as.character(x))

rp_chip <- function(variant, label) {
  stopifnot(variant %in% c("pass", "fail", "neutral", "caution", "void"))
  glyph <- c(pass = "✓", fail = "✕", neutral = "○",
             caution = "!", void = "—")[[variant]]
  paste0('<span class="ca-chip ca-chip--', variant, '"><span class="ca-chip__g">',
         glyph, '</span>', esc(label), '</span>')
}

rp_prov <- function(txt) paste0('<p class="ca-provenance">', esc(txt), '</p>')

rp_tech <- function(body, title = "The package's own words") {
  paste0('<details class="ca-tech"><summary>', esc(title),
         '</summary><div class="ca-tech__body"><p>', esc(body), '</p></div></details>')
}

rp_card <- function(title, body_html, chip = NULL, lede = NULL, foot = NULL) {
  paste0(
    '<section class="ca-card"><header class="ca-card__head">',
    '<h3 class="ca-card__title">', esc(title), '</h3>',
    if (!is.null(chip)) chip else "",
    '</header>',
    if (!is.null(lede)) paste0('<p class="ca-card__lede">', esc(lede), '</p>') else "",
    '<div class="ca-card__body">', body_html, '</div>',
    if (!is.null(foot)) paste0('<div class="ca-card__foot">', foot, '</div>') else "",
    '</section>'
  )
}

rp_stat_line <- function(value_txt, label = "statistic") {
  paste0('<p class="ca-stat"><span class="ca-stat__value">', esc(value_txt),
         '</span> <span class="ca-stat__label">', esc(label), '</span></p>')
}

rp_kv_table <- function(rows) {
  cells <- vapply(rows, function(r) {
    paste0('<tr><th scope="row">', esc(r[[1]]), '</th><td>', esc(r[[2]]), '</td></tr>')
  }, character(1))
  paste0('<table class="ca-kv">', paste(cells, collapse = ""), '</table>')
}

# ---- the six descriptive numbers (helpers.R F.1; permitted app-side) -------
rp_summary <- function(prepared) {
  if (is.null(prepared) || !is.data.frame(prepared) || nrow(prepared) == 0L ||
      is.null(prepared$Y) || is.null(prepared$D)) {
    return(list(n = 0L, events = 0L, censored_pct = NA_real_,
                median_followup = NA_real_, max_followup = NA_real_,
                last_event_time = NA_real_))
  }
  y <- as.numeric(prepared$Y); d <- as.integer(prepared$D)
  ev <- sum(d == 1L, na.rm = TRUE)
  list(
    n = length(y),
    events = ev,
    censored_pct = 100 * (1 - mean(d == 1L, na.rm = TRUE)),
    median_followup = stats::median(y, na.rm = TRUE),
    max_followup = max(y, na.rm = TRUE),
    last_event_time = if (ev > 0L) max(y[d == 1L], na.rm = TRUE) else NA_real_
  )
}

# ---- the one permitted closed form (helpers.R F.4) ------------------------
# qn.test() returns method, statistic and interpretation only: the threshold
# 1 - alpha^(1/n) sits inside the interpretation sentence and in no field, so
# the app (and this report) evaluate it. It is a closed form in alpha and the
# sample size alone. For qn, LARGER is better.
rp_qn_threshold <- function(n, alpha = 0.05) {
  n <- suppressWarnings(as.numeric(n)); a <- suppressWarnings(as.numeric(alpha))
  if (length(n) != 1L || !is.finite(n) || n < 1) return(NA_real_)
  if (length(a) != 1L || !is.finite(a) || a <= 0 || a >= 1) return(NA_real_)
  1 - a^(1 / n)
}

# alpha in force: read off the package's own mz/shen objects, never invented.
ALPHA <- 0.05
if (HAVE_TESTS) {
  a <- rp_stat(TESTS$mz$alpha)
  if (is.na(a)) a <- rp_stat(TESTS$shen$alpha)
  if (!is.na(a)) ALPHA <- a
}
N_PREP <- if (HAVE_PREP) nrow(P_PREPARED) else NA_integer_

# ---- the three follow-up tests, as chips -----------------------------------
# Directions are the contract's: Maller-Zhou and Shen are SMALLER IS BETTER
# against alpha; qn is LARGER IS BETTER against 1 - alpha^(1/n). Getting qn's
# direction backwards is a blocking bug. Note that mz and qn are algebraically
# the same test - mz = (1 - qn)^n exactly - so they can never disagree and are
# never independent confirmation of one another.
rp_followup <- function() {
  if (!HAVE_TESTS) return(NULL)
  a_mz <- rp_stat(TESTS$mz$alpha);   if (is.na(a_mz)) a_mz <- ALPHA
  a_sh <- rp_stat(TESTS$shen$alpha); if (is.na(a_sh)) a_sh <- ALPHA
  qn_thr <- rp_qn_threshold(N_PREP, ALPHA)

  mz <- rp_stat(TESTS$mz$statistic)
  qn <- rp_stat(TESTS$qn$statistic)
  sh <- rp_stat(TESTS$shen$statistic)

  list(
    list(name = "Maller–Zhou", stat = mz, threshold = a_mz, computable = !is.na(mz),
         sufficient = !is.na(mz) && mz < a_mz),
    list(name = "qn", stat = qn, threshold = qn_thr, computable = !is.na(qn),
         sufficient = !is.na(qn) && !is.na(qn_thr) && qn > qn_thr),
    list(name = "Shen", stat = sh, threshold = a_sh, computable = !is.na(sh),
         sufficient = !is.na(sh) && sh < a_sh)
  )
}
FU <- rp_followup()

# The fixed decision-string to banner-variant lookup. An unrecognised string is
# a verdict this report cannot name, so it must not style it either.
RP_REC_CLASS <- c(
  "Cure model appropriate"                   = "ca-rec--appropriate",
  "Cure model not supported"                 = "ca-rec--unsupported",
  "Follow-up insufficient for cure modeling" = "ca-rec--insufficient"
)
rp_rec_class <- function(decision) {
  if (is.null(decision) || length(decision) != 1L || is.na(decision) ||
      !(decision %in% names(RP_REC_CLASS))) return(NA_character_)
  unname(RP_REC_CLASS[[decision]])
}

# ---- the not-computable wording, shared by three cards (contract J.10/J.11)
RP_NOT_COMPUTABLE_1 <- "Cannot be computed — the longest observed time is an event, so there is no plateau to test."
RP_NOT_COMPUTABLE_2 <- "This is a property of the data, not an error. All three follow-up tests need follow-up to extend past the last event."

rp_followup_card <- function(title, obj, stat, threshold, thr_text, dir_text,
                             sufficient, provenance) {
  if (is.na(stat)) {
    body <- paste0(
      rp_stat_line(rp_dash()),
      '<p class="ca-void-line"><strong>', esc(RP_NOT_COMPUTABLE_1), '</strong></p>',
      '<p>', esc(RP_NOT_COMPUTABLE_2), '</p>',
      rp_tech(obj$interpretation),
      rp_prov(provenance)
    )
    return(rp_card(title, body, chip = rp_chip("void", "Cannot be computed")))
  }
  body <- paste0(
    rp_stat_line(rp_num(stat)),
    '<p class="ca-threshold"><span class="ca-threshold__value">', esc(thr_text),
    '</span> <span class="ca-threshold__dir">', esc(dir_text), '</span></p>',
    rp_tech(obj$interpretation),
    rp_prov(provenance)
  )
  rp_card(
    title, body,
    chip = if (isTRUE(sufficient)) rp_chip("pass", "Follow-up looks sufficient")
           else rp_chip("fail", "Not sufficient")
  )
}
```

<style type="text/css">
/* FIXPASS: inlined, never linked from app/www/app.css, so the output HTML is
   one self-contained file that opens offline. The tokens are the app's own. */
:root {
  --ca-surface:#FBFCFC; --ca-canvas:#EEF2F3; --ca-sunk:#F4F7F8;
  --ca-ink:#11191C; --ca-ink2:#414D52; --ca-ink3:#667378;
  --ca-rule:#DBE2E4; --ca-rule2:#C3CDD0; --ca-accent:#0A5670;
  --ca-pass-ink:#175A2B; --ca-pass-bg:#E6F3E9; --ca-pass-line:#B4D7BE;
  --ca-fail-ink:#8E2C20; --ca-fail-bg:#FBEAE6; --ca-fail-line:#EDC2B8;
  --ca-caution-ink:#7A4E05; --ca-caution-bg:#FBF1DE; --ca-caution-line:#E8D0A2;
  --ca-neutral-ink:#0A5670; --ca-neutral-bg:#E6F0F4; --ca-neutral-line:#BCD5E0;
  --ca-void-ink:#4B565A; --ca-void-bg:#ECEFF0; --ca-void-line:#CBD3D5;
  --ca-sans: system-ui, -apple-system, BlinkMacSystemFont, "Segoe UI", Roboto, "Helvetica Neue", Arial, sans-serif;
  --ca-serif: Charter, "Bitstream Charter", "Sitka Text", Cambria, "Iowan Old Style", "Palatino Linotype", Georgia, serif;
  --ca-mono: ui-monospace, SFMono-Regular, "SF Mono", Menlo, Consolas, "Liberation Mono", monospace;
}
body { font-family: var(--ca-sans); font-size: 16px; line-height: 1.55;
       color: var(--ca-ink); background: var(--ca-canvas);
       max-width: 62rem; margin: 0 auto; padding: 2rem 1.25rem 4rem; }
h1, h2, h3, h4 { font-family: var(--ca-serif); font-weight: 600; color: var(--ca-ink);
                 line-height: 1.25; }
h1 { font-size: 2rem; margin: 0 0 .25rem; }
h2 { font-size: 1.5rem; margin: 2.25rem 0 .75rem; padding-bottom: .4rem;
     border-bottom: 2px solid var(--ca-rule2); }
h3 { font-size: 1.2rem; margin: 0; }
h4 { font-size: 1.05rem; margin: 1.5rem 0 .5rem; }
a { color: var(--ca-accent); }
code, .ca-mono { font-family: var(--ca-mono); font-size: .9em; }
#TOC { background: var(--ca-surface); border: 1px solid var(--ca-rule);
       border-radius: 10px; padding: .75rem 1rem; margin: 1.5rem 0; }
#TOC ul { margin: .25rem 0; padding-left: 1.1rem; }

.ca-card { background: var(--ca-surface); border: 1px solid var(--ca-rule);
           border-radius: 10px; padding: 1rem 1.15rem; margin: 0 0 1rem; }
.ca-card__head { display: flex; flex-wrap: wrap; align-items: baseline;
                 justify-content: space-between; gap: .5rem; margin-bottom: .5rem; }
.ca-card__lede { color: var(--ca-ink2); margin: 0 0 .6rem; }
.ca-card__body > :first-child { margin-top: 0; }
.ca-card__body > :last-child { margin-bottom: 0; }
.ca-card__foot { margin-top: .6rem; border-top: 1px solid var(--ca-rule);
                 padding-top: .5rem; }

.ca-chip { display: inline-block; border: 1px solid; border-radius: 999px;
           padding: .15rem .65rem; font-size: .9rem; font-weight: 600; }
.ca-chip__g { margin-right: .4rem; font-weight: 700; }
.ca-chip--pass    { color: var(--ca-pass-ink);    background: var(--ca-pass-bg);    border-color: var(--ca-pass-line); }
.ca-chip--fail    { color: var(--ca-fail-ink);    background: var(--ca-fail-bg);    border-color: var(--ca-fail-line); }
.ca-chip--caution { color: var(--ca-caution-ink); background: var(--ca-caution-bg); border-color: var(--ca-caution-line); }
.ca-chip--neutral { color: var(--ca-neutral-ink); background: var(--ca-neutral-bg); border-color: var(--ca-neutral-line); }
.ca-chip--void    { color: var(--ca-void-ink);    background: var(--ca-void-bg);    border-color: var(--ca-void-line); font-weight: 500; }

.ca-stat { margin: .3rem 0 .4rem; }
.ca-stat__value { font-size: 1.7rem; font-weight: 700; font-family: var(--ca-mono); }
.ca-stat__label { color: var(--ca-ink3); font-size: .85rem; text-transform: uppercase;
                  letter-spacing: .04em; }
.ca-threshold { margin: 0 0 .5rem; color: var(--ca-ink2); }
.ca-threshold__value { font-family: var(--ca-mono); }
.ca-threshold__dir { color: var(--ca-ink3); font-size: .9rem; }
.ca-void-line { color: var(--ca-void-ink); }

.ca-provenance { font-family: var(--ca-mono); font-size: .8rem; color: var(--ca-ink3);
                 margin: .5rem 0 0; overflow-wrap: anywhere; }
.ca-tech { margin: .5rem 0 0; border: 1px solid var(--ca-rule); border-radius: 6px;
           background: var(--ca-sunk); }
.ca-tech > summary { cursor: pointer; padding: .4rem .7rem; font-size: .9rem;
                     color: var(--ca-ink2); }
.ca-tech__body { padding: 0 .7rem .6rem; }
.ca-tech__body p { margin: 0; }

.ca-statrow { display: flex; flex-wrap: wrap; gap: .75rem; }
.ca-statrow__cell { flex: 1 1 8.5rem; background: var(--ca-surface);
                    border: 1px solid var(--ca-rule); border-radius: 8px; padding: .6rem .75rem; }
.ca-statrow__label { display: block; font-size: .75rem; text-transform: uppercase;
                     letter-spacing: .05em; color: var(--ca-ink3); }
.ca-statrow__value { display: block; font-size: 1.45rem; font-weight: 700;
                     font-family: var(--ca-mono); }
.ca-statrow__note { display: block; font-size: .78rem; color: var(--ca-ink3); }

table.ca-kv { border-collapse: collapse; width: 100%; }
table.ca-kv th, table.ca-kv td { border-bottom: 1px solid var(--ca-rule);
                                 padding: .35rem .5rem; text-align: left; vertical-align: top; }
table.ca-kv th { width: 34%; font-weight: 600; color: var(--ca-ink2); }

.ca-tablewrap { overflow-x: auto; }
table.ca-aic { border-collapse: collapse; width: 100%; font-size: .9rem;
               background: var(--ca-surface); }
table.ca-aic th, table.ca-aic td { border: 1px solid var(--ca-rule); padding: .35rem .5rem;
                                   text-align: left; vertical-align: top; }
table.ca-aic thead th { background: var(--ca-sunk); }
table.ca-aic td.num { font-family: var(--ca-mono); text-align: right; white-space: nowrap; }
table.ca-aic td.err { color: var(--ca-fail-ink); }
.ca-badge { display: inline-block; border: 1px solid var(--ca-rule2); border-radius: 4px;
            padding: 0 .35rem; font-size: .8rem; }

.ca-note { border: 1px solid var(--ca-rule); border-left-width: 5px; border-radius: 8px;
           background: var(--ca-surface); padding: .8rem 1rem; margin: 1rem 0; }
.ca-note__title { font-family: var(--ca-serif); font-weight: 600; margin: 0 0 .35rem; }
.ca-note--info { border-left-color: var(--ca-neutral-ink); }
.ca-note--warning { border-left-color: var(--ca-caution-ink); background: var(--ca-caution-bg); }
.ca-note--interpret { border-left-color: var(--ca-accent); }

.ca-steps { display: flex; flex-direction: column; gap: .75rem; }
.ca-step { display: flex; gap: .9rem; background: var(--ca-surface);
           border: 1px solid var(--ca-rule); border-left-width: 5px; border-radius: 10px;
           padding: .9rem 1rem; }
.ca-step[data-ca-state="pass"] { border-left-color: var(--ca-pass-ink); }
.ca-step[data-ca-state="fail"] { border-left-color: var(--ca-fail-ink); }
.ca-step[data-ca-state="pending"] { border-left-color: var(--ca-caution-ink); }
.ca-step[data-ca-state="void"] { border-left-color: var(--ca-rule2); }
.ca-step__marker { font-size: 1.6rem; font-weight: 700; color: var(--ca-ink3); line-height: 1; }
.ca-step__title { font-family: var(--ca-serif); font-size: 1.1rem; margin: 0 0 .35rem; }
.ca-step__body p { margin: .4rem 0 0; }
.ca-step__body blockquote { margin: .5rem 0 0; padding-left: .8rem;
                            border-left: 3px solid var(--ca-rule2); color: var(--ca-ink2); }

.ca-rec { border: 1px solid var(--ca-rec-line, var(--ca-rule)); border-radius: 10px;
          background: var(--ca-rec-bg, var(--ca-surface)); padding: 1rem 1.15rem; margin: 1rem 0; }
.ca-rec--appropriate { --ca-rec-line: var(--ca-pass-line); --ca-rec-bg: var(--ca-pass-bg); --ca-rec-ink: var(--ca-pass-ink); }
.ca-rec--unsupported { --ca-rec-line: var(--ca-fail-line); --ca-rec-bg: var(--ca-fail-bg); --ca-rec-ink: var(--ca-fail-ink); }
.ca-rec--insufficient { --ca-rec-line: var(--ca-caution-line); --ca-rec-bg: var(--ca-caution-bg); --ca-rec-ink: var(--ca-caution-ink); }
.ca-rec__decision { font-family: var(--ca-serif); font-size: 1.65rem; font-weight: 700;
                    color: var(--ca-rec-ink, var(--ca-ink)); margin-bottom: .6rem; }
.ca-rec__conditions { display: flex; flex-wrap: wrap; gap: .5rem; margin-bottom: .6rem; }
.ca-rec__cond { font-family: var(--ca-mono); font-size: .95rem; background: var(--ca-surface);
                border: 1px solid var(--ca-rule); border-radius: 6px; padding: .25rem .6rem; }
.ca-rec__cond[data-ca-met="true"]::before { content: "\2713 "; color: var(--ca-pass-ink); font-weight: 700; }
.ca-rec__cond[data-ca-met="false"]::before { content: "\2715 "; color: var(--ca-fail-ink); font-weight: 700; }
.ca-key { font-weight: 600; }
.ca-lede { color: var(--ca-ink2); }
.ca-foot { color: var(--ca-ink3); font-size: .85rem; border-top: 1px solid var(--ca-rule);
           margin-top: 2.5rem; padding-top: .75rem; }
pre { background: var(--ca-sunk); border: 1px solid var(--ca-rule); border-radius: 6px;
      padding: .75rem; overflow-x: auto; font-size: .82rem; }
@media print { body { background: #fff; } .ca-tech { display: none; } }
</style>

```{r title-block, results='asis'}
lbl <- if (!is.null(params$label) && nzchar(as.character(params$label)[1])) {
  as.character(params$label)[1]
} else "no dataset supplied"
cat('<p class="ca-lede">Dataset: <strong>', esc(lbl), '</strong> · generated ',
    esc(format(Sys.time(), "%Y-%m-%d %H:%M")),
    ' by cureAssessApp, a front end to the <span class="ca-mono">cureAssess</span> R package.</p>\n',
    sep = "")
cat('<div class="ca-note ca-note--info"><p class="ca-note__title">What this report is</p>',
    '<p>This app assesses whether a cure model is <em>appropriate</em>. It does not fit your final ',
    'analysis model, and it does not produce treatment-effect estimates.</p>',
    '<p>Nothing in this document was computed by the app. Every statistic below is a field of an ',
    'object returned by an exported <span class="ca-mono">cureAssess</span> function, and each one ',
    'names its call and field.</p></div>\n', sep = "")
```

```{r nothing-supplied, results='asis', eval=!HAVE_ANY}
cat('<div class="ca-note ca-note--warning"><p class="ca-note__title">No assessment supplied</p>',
    '<p>This report was knitted with its default parameters, so there is no dataset and no ',
    'assessment to describe. Every section below needs the objects the app passes in: the prepared ',
    'data frame from <span class="ca-mono">prepare.surv.data()</span>, the fit from ',
    '<span class="ca-mono">model.fitting()</span>, and the assessment from ',
    '<span class="ca-mono">cure.appropriateness(run_tests = "yes")</span>.</p>',
    '<p>To produce a real report, download it from the Conclusion step of the app after running an ',
    'assessment, or knit this file with those objects passed as parameters.</p></div>\n', sep = "")
```

## Dataset and provenance

```{r provenance, results='asis'}
src <- params$source
src_txt <- if (identical(src, "builtin")) {
  "Built-in example shipped with the app"
} else if (identical(src, "upload")) {
  "Uploaded by the user"
} else {
  rp_dash()
}
scale_txt <- if (!is.null(P_MAP$time_scale)) {
  if (identical(P_MAP$time_scale, "days_to_years")) {
    "days_to_years — times were divided by 365.25 by prepare.surv.data()"
  } else {
    "none — times were used as supplied"
  }
} else rp_dash()

rows <- list(
  c("Dataset", lbl),
  c("Provenance", src_txt),
  c("Time column", if (!is.null(P_MAP$time)) P_MAP$time else rp_dash()),
  c("Status column", if (!is.null(P_MAP$status)) P_MAP$status else rp_dash()),
  c("Value meaning “event”", if (!is.null(P_MAP$event_level)) P_MAP$event_level else rp_dash()),
  c("Time scale", scale_txt)
)
cat(rp_card(
  "How this dataset was read",
  paste0(
    rp_kv_table(rows),
    '<p class="ca-lede">The status column was recoded to 0/1 by the app before the package saw it: ',
    'a row counts as an event when the status value equals the level named above. Rows with a ',
    'missing time or status were dropped, so every n in this report is the analysed n.</p>',
    rp_prov("prepare.surv.data(data, time, status, time_scale) -> the prepared frame")
  )
), "\n")
```

## Data summary

```{r data-summary, results='asis'}
if (!HAVE_PREP) {
  cat('<p class="ca-lede">No prepared dataset was supplied, so there is nothing to summarise.</p>\n')
} else {
  s <- rp_summary(P_PREPARED)
  cells <- list(
    c("n", format(s$n, big.mark = ","), "rows analysed, after dropping"),
    c("Events", format(s$events, big.mark = ","), "D = 1"),
    c("Censored", rp_pct(s$censored_pct), "share of rows with D = 0"),
    c("Median follow-up", rp_num(s$median_followup, 3), "median of observed times, all subjects"),
    c("Max follow-up", rp_num(s$max_followup, 3), "largest observed time"),
    c("Last event", rp_num(s$last_event_time, 3), "largest time with D = 1")
  )
  html <- paste0(vapply(cells, function(c3) paste0(
    '<div class="ca-statrow__cell"><span class="ca-statrow__label">', esc(c3[1]),
    '</span><span class="ca-statrow__value">', esc(c3[2]),
    '</span><span class="ca-statrow__note">', esc(c3[3]), '</span></div>'
  ), character(1)), collapse = "")
  cat('<div class="ca-statrow">', html, '</div>\n', sep = "")
  cat('<p class="ca-lede">These six are descriptive counts on the prepared <span class="ca-mono">Y</span> ',
      'and <span class="ca-mono">D</span> columns, not cure-model statistics. They are the only ',
      'numbers in this report the app computed itself.</p>\n', sep = "")
}
```

## Kaplan-Meier curve

```{r km-plot, fig.alt="Kaplan-Meier survival curve with a number-at-risk table underneath", results='asis', eval=HAVE_KM}
cat('<p class="ca-lede">The package’s own Kaplan-Meier plot, with its risk table, exactly as the ',
    'app shows it.</p>\n', sep = "")
```

```{r km-draw, eval=HAVE_KM}
ok <- tryCatch({
  if (requireNamespace("survminer", quietly = TRUE)) invisible(NULL)
  print(P_FIT$kmplot)
  TRUE
}, error = function(e) {
  message("KM plot could not be drawn: ", conditionMessage(e))
  FALSE
})
```

```{r km-prov, results='asis', eval=HAVE_KM}
cat(rp_prov("model.fitting(plot_km = TRUE) -> $kmplot"), "\n")
cat('<p class="ca-lede">A plateau is not proof. A Kaplan-Meier curve goes flat for two different ',
    'reasons that look identical on the plot: the uncured have all had their events, or there is ',
    'nobody left to watch. Heavy censoring alone manufactures a convincing plateau in data with a ',
    'true cure fraction of exactly zero. That is what the diagnostics below are for.</p>\n', sep = "")
```

```{r km-missing, results='asis', eval=!HAVE_KM}
cat('<p class="ca-lede">No Kaplan-Meier plot was supplied. The app passes ',
    '<span class="ca-mono">model.fitting(plot_km = TRUE)$kmplot</span>; without it there is no ',
    'curve to draw.</p>\n', sep = "")
```

## Model screening (AIC)

```{r aic-table, results='asis'}
if (!HAVE_AIC) {
  cat('<p class="ca-lede">No assessment was supplied, so there is no AIC table.</p>\n')
} else {
  tb <- P_ASSESS$screening$aic_table
  n_models <- nrow(tb)
  set_txt <- if (isTRUE(params$include_lognormal)) {
    paste0("Candidate set: ", n_models, " models, with include_lognormal = TRUE — the lognormal ",
           "models were fitted. The lognormal tail can change both the selected model and the ",
           "RECeUS conclusion.")
  } else {
    paste0("Candidate set: ", n_models, " models, with include_lognormal = FALSE — the lognormal ",
           "models were not fitted. This is the default and matches the tutorial’s four ",
           "distributions.")
  }
  cat('<p class="ca-lede">', esc(set_txt), '</p>\n', sep = "")

  hdr <- paste0('<tr><th>model</th><th>model_type</th><th>AIC</th>',
                '<th>parameter_estimates</th><th>error</th></tr>')
  body <- vapply(seq_len(n_models), function(i) {
    paste0(
      '<tr><td><span class="ca-mono">', esc(tb$model[i]), '</span></td>',
      '<td><span class="ca-badge">', esc(tb$model_type[i]), '</span></td>',
      '<td class="num">', esc(rp_num(tb$AIC[i], 2)), '</td>',
      '<td><span class="ca-mono">', esc(tb$parameter_estimates[i]), '</span></td>',
      '<td class="err">', esc(tb$error[i]), '</td></tr>'
    )
  }, character(1))
  cat('<div class="ca-tablewrap"><table class="ca-aic"><thead>', hdr, '</thead><tbody>',
      paste(body, collapse = ""), '</tbody></table></div>\n', sep = "")
  cat('<p class="ca-lede">The <span class="ca-mono">error</span> column is always shown and failed ',
      'rows are never removed: <span class="ca-mono">model.fitting()</span> does not throw on a ',
      'failed fit, it records the reason there and sets <span class="ca-mono">AIC = NA</span>. ',
      'Hiding the column would turn a silent failure into an invisible one.</p>\n', sep = "")

  best <- P_ASSESS$screening$best_model
  btyp <- P_ASSESS$screening$best_model_type
  cat(rp_card(
    "Smallest AIC",
    paste0(
      '<p><strong>', esc(if (length(best) == 1L && !is.na(best)) best else rp_dash()),
      '</strong> <span class="ca-badge">',
      esc(if (length(btyp) == 1L && !is.na(btyp)) btyp else rp_dash()), '</span></p>',
      '<blockquote><p>', esc(P_ASSESS$screening$initial_decision), '</p></blockquote>',
      '<p class="ca-lede">AIC is initial support only. It asks how well a model describes the data ',
      'you have; identifiability asks whether the data contain enough follow-up to pin a cure ',
      'fraction down. Only the second question is decided by the diagnostics below.</p>',
      rp_prov("cure.appropriateness() -> $screening$aic_table, $best_model, $initial_decision")
    )
  ), "\n")
}
```

## The five diagnostics

```{r diag-lede, results='asis', eval=HAVE_TESTS}
cat('<p class="ca-lede">All five are shown, including when a non-cure model won on AIC: ',
    '“a non-cure model fits better” and “there is no cure fraction to find” are ',
    'different claims. The RECeUS distribution used was ',
    '<span class="ca-mono">', esc(P_ASSESS$selected_receus_dist), '</span>',
    if (!is.null(P_ASSESS$selected_receus_model) && length(P_ASSESS$selected_receus_model) == 1L &&
        !is.na(P_ASSESS$selected_receus_model)) {
      paste0(', mapped from <span class="ca-mono">', esc(P_ASSESS$selected_receus_model), '</span>')
    } else "",
    '. Maller–Zhou and <span class="ca-mono">qn</span> are the same underlying count presented ',
    'two ways, so they always agree with each other and are never independent confirmation.</p>\n',
    sep = "")
```

```{r diag-cards, results='asis', eval=HAVE_TESTS}
qn_thr <- rp_qn_threshold(N_PREP, ALPHA)

cat(rp_followup_card(
  "Maller–Zhou test (1994)", TESTS$mz, FU[[1]]$stat, FU[[1]]$threshold,
  paste0("Threshold: α = ", rp_num(FU[[1]]$threshold, 3)),
  "smaller is better — below α supports sufficient follow-up",
  FU[[1]]$sufficient,
  "cure.appropriateness() -> $tests$mz$statistic, $alpha"
), "\n")

cat(rp_followup_card(
  "qn statistic (Maller & Zhou 1996)", TESTS$qn, FU[[2]]$stat, FU[[2]]$threshold,
  paste0("Threshold: 1 - α^(1/n) = ", rp_num(qn_thr, 6), "  (computed by this app)"),
  "larger is better — above the threshold supports sufficient follow-up",
  FU[[2]]$sufficient,
  "cure.appropriateness() -> $tests$qn$statistic; threshold computed by the app from α and n"
), "\n")

cat(rp_followup_card(
  "Shen test (2000)", TESTS$shen, FU[[3]]$stat, FU[[3]]$threshold,
  paste0("Threshold: α = ", rp_num(FU[[3]]$threshold, 3)),
  "smaller is better — below α supports sufficient follow-up; Shen uses a narrower window than Maller–Zhou and is the stricter of the two by design",
  FU[[3]]$sufficient,
  "cure.appropriateness() -> $tests$shen$statistic, $alpha"
), "\n")

# Immune summary: descriptive, and never a pass/fail chip, despite the name.
im <- TESTS$immune
cat(rp_card(
  "Maller–Zhou immune summary (1996)",
  paste0(
    rp_kv_table(list(
      c("Event probability by the end of follow-up (p_hat)", rp_num(im$p_hat)),
      c("Censoring proportion (p_cens)", rp_num(im$p_cens)),
      c("Last observation", rp_num(im$last_observation)),
      c("Last observation censored?",
        if (isTRUE(im$last_observation_censored)) "TRUE — follow-up outlived the last event"
        else "FALSE — the study stopped while events were still happening")
    )),
    '<p class="ca-lede">There is no threshold and no null hypothesis here. This is a descriptive ',
    'summary, not a test, and it never carries a pass or fail.</p>',
    rp_tech(im$interpretation),
    rp_prov("cure.appropriateness() -> $tests$immune$p_hat, $p_cens, $last_observation, $last_observation_censored")
  ),
  chip = rp_chip("neutral", "Descriptive, not a test")
), "\n")

# RECeUS.
rc <- TESTS$receus
pi_hat <- rp_stat(rc$pi_hat); r_hat <- rp_stat(rc$r_hat); tau <- rp_stat(rc$tau)
if (is.na(pi_hat) && is.na(r_hat)) {
  cat(rp_card(
    "RECeUS (Selukar & Othus, 2023)",
    paste0(
      rp_stat_line(rp_dash(), "pi_hat"), rp_stat_line(rp_dash(), "r_hat"),
      '<p class="ca-void-line"><strong>Cannot be computed — the model fit behind RECeUS did not ',
      'converge for this dataset.</strong></p>',
      rp_tech(rc$interpretation),
      rp_prov("cure.appropriateness() -> $tests$receus$pi_hat, $r_hat")
    ),
    chip = rp_chip("void", "Cannot be computed")
  ), "\n")
} else {
  chip_variant <- if (identical(rc$decision, "Cure model appropriate")) "pass" else "fail"
  cat(rp_card(
    "RECeUS (Selukar & Othus, 2023)",
    paste0(
      '<div class="ca-rec__conditions">',
      '<div class="ca-rec__cond" data-ca-met="',
      if (isTRUE(rc$cure_fraction_condition)) "true" else "false", '">π̂ = ',
      esc(rp_num(pi_hat)), ' &gt; 0.025</div>',
      '<div class="ca-rec__cond" data-ca-met="',
      if (isTRUE(rc$followup_condition)) "true" else "false", '">r̂ = ',
      esc(rp_num(r_hat)), ' &lt; 0.05</div></div>',
      '<p>Evaluated at τ = ', esc(rp_num(tau)), ', the largest observed time. Distribution: ',
      '<span class="ca-mono">', esc(rc$dist), '</span>.</p>',
      '<p class="ca-lede">The rule is a conjunction: both conditions must hold. r̂ is the share ',
      'of uncured subjects still censored at the end of follow-up, and it asks the identifiability ',
      'question most directly.</p>',
      rp_tech(rc$interpretation),
      rp_prov("cure.appropriateness() -> $tests$receus$pi_hat, $r_hat, $tau, $decision")
    ),
    chip = rp_chip(chip_variant, rc$decision)
  ), "\n")
}
```

```{r diag-not-run, results='asis', eval=(HAVE_ASSESS && !HAVE_TESTS)}
cat('<h3>The diagnostics were not run</h3>\n')
cat('<blockquote><p>', esc(P_ASSESS$tests_reason), '</p></blockquote>\n', sep = "")
cat('<p class="ca-lede">No diagnostic cards are shown, because there is nothing to show. Steps ',
    '② and ③ of the verdict below are not fabricated either.</p>\n', sep = "")
```

```{r diag-none, results='asis', eval=!HAVE_ASSESS}
cat('<p class="ca-lede">No assessment was supplied, so there are no diagnostics.</p>\n')
```

## The three-step check, in the published order

```{r steps-lede, results='asis'}
cat('<p class="ca-lede">Failing any one step means a cure model is inappropriate. This is a ',
    'conjunction, not a score.</p>\n', sep = "")
```

```{r step-strip, results='asis'}
step <- function(marker, title, state_word, chip, body_html) {
  paste0('<div class="ca-step" data-ca-state="', state_word, '">',
         '<div class="ca-step__marker">', marker, '</div>',
         '<div class="ca-step__body"><h3 class="ca-step__title">', esc(title), '</h3>',
         chip, body_html, '</div></div>')
}

confirmed <- isTRUE(params$expert_confirmed)
step1 <- step(
  "①", "Expert judgment",
  if (confirmed) "pass" else "pending",
  if (confirmed) rp_chip("pass", "Confirmed by user") else rp_chip("neutral", "Still open"),
  paste0('<p>Software cannot judge clinical plausibility. Is there a clinical reason to expect that ',
         'some patients in this population are cured? This step is a conversation with a clinician, ',
         'and this app will never answer it for you.</p>',
         if (confirmed) {
           '<p class="ca-lede">The user confirmed this step in the app before downloading this report.</p>'
         } else {
           '<p class="ca-lede">This step was <strong>not</strong> confirmed in the app. Steps 2 and 3 below still report their findings, but a cure model is only appropriate if all three steps pass.</p>'
         })
)

im <- if (HAVE_TESTS) TESTS$immune else NULL
step2 <- if (is.null(im)) {
  step("②", "Visual assessment", "void", rp_chip("neutral", "Not available"),
       '<p>This is a visual aid, not a test.</p><p class="ca-lede">No assessment was supplied.</p>')
} else {
  step("②", "Visual assessment", "void",
       if (isTRUE(im$last_observation_censored)) rp_chip("neutral", "Plateau possible")
       else rp_chip("neutral", "No clear plateau"),
       paste0('<p>This is a visual aid, not a test.</p><blockquote><p>', esc(im$interpretation),
              '</p></blockquote>',
              rp_prov("cure.appropriateness() -> $tests$immune$last_observation_censored")))
}

rc <- if (HAVE_TESTS) TESTS$receus else NULL
step3 <- if (is.null(rc) || is.null(rc$decision)) {
  step("③", "Quantitative assessment", "void", rp_chip("neutral", "Not available"),
       '<p class="ca-lede">No assessment was supplied.</p>')
} else {
  step("③", "Quantitative assessment",
       if (identical(rc$decision, "Cure model appropriate")) "pass" else "fail",
       rp_chip(if (identical(rc$decision, "Cure model appropriate")) "pass" else "fail", rc$decision),
       rp_prov("cure.appropriateness() -> $tests$receus$decision"))
}

cat('<div class="ca-steps">', step1, step2, step3, '</div>\n', sep = "")
```

```{r recommendation, results='asis', eval=HAVE_TESTS}
rc <- TESTS$receus
variant <- rp_rec_class(rc$decision)
pi_hat <- rp_stat(rc$pi_hat); r_hat <- rp_stat(rc$r_hat); tau <- rp_stat(rc$tau)

m <- sum(vapply(FU, function(x) isTRUE(x$computable), logical(1)))
k <- sum(vapply(FU, function(x) isTRUE(x$sufficient), logical(1)))
tally <- paste0("Of the three follow-up tests, ", k, " of ", m,
                " computable tests indicated sufficient follow-up")
tally <- if (m < 3L) {
  paste0(tally, "; the other ", 3L - m, " could not be computed on this dataset.")
} else {
  paste0(tally, ".")
}

cat('<div class="ca-rec ', if (is.na(variant)) "" else variant, '">',
    '<div class="ca-rec__decision">', esc(rc$decision), '</div>',
    '<div class="ca-rec__conditions">',
    '<div class="ca-rec__cond" data-ca-met="',
    if (isTRUE(rc$cure_fraction_condition)) "true" else "false", '">π̂ = ',
    esc(rp_num(pi_hat)), ' &gt; 0.025</div>',
    '<div class="ca-rec__cond" data-ca-met="',
    if (isTRUE(rc$followup_condition)) "true" else "false", '">r̂ = ',
    esc(rp_num(r_hat)), ' &lt; 0.05</div></div>',
    '<p>Evaluated at τ = ', esc(rp_num(tau)), ', the largest observed time.</p>',
    '<p class="ca-key">', esc(tally), '</p>',
    '<p class="ca-lede">That line counts chips already shown above. It is not a composite score, ',
    'the tests are not weighted, and a majority does not decide anything.</p>',
    rp_prov("cure.appropriateness() -> $tests$receus$pi_hat, $r_hat, $decision"),
    '</div>\n', sep = "")
```

## What the package says

```{r package-says, results='asis', eval=HAVE_ASSESS}
cat(rp_card(
  "The package's own summary",
  paste0(
    '<blockquote><p>', esc(P_ASSESS$final_recommendation), '</p></blockquote>',
    '<p class="ca-lede">This sentence summarises the model comparison only — it does not mention ',
    'RECeUS, π̂ or r̂. The RECeUS recommendation is step ③ above.</p>',
    '<h4>Why the diagnostics were or were not run</h4>',
    '<blockquote><p>', esc(P_ASSESS$tests_reason), '</p></blockquote>',
    rp_prov("cure.appropriateness() -> $final_recommendation, $tests_reason")
  )
), "\n")
```

```{r package-says-none, results='asis', eval=!HAVE_ASSESS}
cat('<p class="ca-lede">No assessment was supplied, so the package said nothing.</p>\n')
```

## The diagnostics can disagree

```{r disagree-lead, results='asis', eval=HAVE_TESTS}
rc <- TESTS$receus
rows <- FU
if (!is.na(rp_stat(rc$r_hat))) {
  rows <- c(rows, list(list(name = "RECeUS", computable = TRUE,
                            sufficient = isTRUE(rc$followup_condition))))
}
computable <- Filter(function(x) isTRUE(x$computable), rows)
yes <- vapply(Filter(function(x) isTRUE(x$sufficient), computable), function(x) x$name, character(1))
no  <- vapply(Filter(function(x) !isTRUE(x$sufficient), computable), function(x) x$name, character(1))
if (length(yes) && length(no)) {
  cat('<p><strong>On this dataset the diagnostics disagree.</strong> ',
      esc(paste(yes, collapse = ", ")), ' indicate sufficient follow-up; ',
      esc(paste(no, collapse = ", ")), ' do not. This is expected — read on.</p>\n', sep = "")
}
```

<p><strong>Disagreement is information, not a malfunction.</strong> These five diagnostics ask
overlapping but different questions, and they were published by different authors to catch different
failures. Do not count them and take the majority &mdash; they are not votes. Maller&ndash;Zhou and
<span class="ca-mono">qn</span> are the same underlying count presented two ways, so they always
agree with each other and are never independent confirmation of one another. Shen uses a narrower
window precisely because Maller&ndash;Zhou can over-declare sufficiency, so Maller&ndash;Zhou passing
while Shen fails is Shen working as designed. RECeUS asks the identifiability question most directly,
through r&#770;. When they split: check how close each statistic sits to its threshold, prefer the
stricter test when the pattern is the known one, weigh r&#770;, and look back at the curve.</p>

<p class="ca-lede">The thresholds 0.05, 0.025 and the r&#770; cut-off are conventions rather than
laws. A dataset landing just the wrong side of one is telling you the answer is genuinely uncertain,
not that it is no.</p>

```{r options-panel, results='asis', eval=HAVE_TESTS}
if (identical(TESTS$receus$decision, "Follow-up insufficient for cure modeling")) {
  cat('<div class="ca-note ca-note--warning">',
      '<p class="ca-note__title">The diagnostics say follow-up is insufficient. What are my options?</p>',
      '<p>This is not a dead end, and it is not a verdict on your data quality. It means the ',
      'follow-up you have cannot separate “cured” from “not yet relapsed”, so a ',
      'cure model would return a number that looks precise and is not identified. Four routes:</p>',
      '<p><strong>1. Fit a non-cure model instead.</strong> The most common and usually the right ',
      'answer. Standard survival models — Cox, Weibull, whatever fits — remain valid and ',
      'interpretable; you simply do not report a cure fraction. You lose a quantity the data could ',
      'not support anyway.</p>',
      '<p><strong>2. Use the extreme-value estimators of Escobar-Bach and Van Keilegom (2019).</strong> ',
      'Non-parametric cure-rate estimation designed specifically for insufficient follow-up, using ',
      'extreme-value theory to extrapolate the tail rather than assuming the plateau is complete. Not ',
      'implemented in cureAssess — you would fit this outside the app.</p>',
      '<p><strong>3. Use Yuen and Musta’s relaxed condition (2024).</strong> A weaker ',
      'sufficient-follow-up requirement than the classical condition used by the tests here, so some ',
      'datasets that fail the classical condition are still workable under theirs. Also outside this ',
      'app.</p>',
      '<p><strong>4. Collect more follow-up.</strong> If the study is ongoing, or a later data cut ',
      'exists, this is the only route that fixes the problem at source rather than working around it. ',
      'Othus et al. refitted cure models on six SWOG trials at an early and a later follow-up time, ',
      'found the mean-survival estimates shifted materially, and found the <strong>direction of the ',
      'shift was not predictable</strong>. There is no post-hoc correction to apply — which is ',
      'why this check happens before you fit, not after.</p></div>\n', sep = "")
}
```

## References and how to cite

The four key references behind the diagnostics in this report:

- Maller RA, Zhou S (1992). Estimating the proportion of immunes in a censored sample. *Biometrika*, 79(4), 731&ndash;739. doi:10.1093/biomet/79.4.731
- Maller RA, Zhou S (1994). Testing for sufficient follow-up and outliers in survival data. *Journal of the American Statistical Association*, 89(428), 1499&ndash;1506. doi:10.1080/01621459.1994.10476889
- Shen P-S (2000). Testing for sufficient follow-up in survival data. *Statistics & Probability Letters*, 49(4), 313&ndash;322. doi:10.1016/S0167-7152(00)00063-8
- Selukar S, Othus M (2023). RECeUS: Ratio estimation of censored uncured subjects, a different approach for assessing cure model appropriateness in studies with long-term survivors. *Statistics in Medicine*, 42(3), 209&ndash;227. doi:10.1002/sim.9610

The workflow this app implements is the Figure 1 workflow of &ldquo;A Tutorial for Evaluating Cure
Model Appropriateness&rdquo; &mdash; Mudunkotuwa, Ghosh, Triplett, Selukar (in preparation).

**How to cite.** Cite the package with `citation("cureAssess")` in R, and cite the tutorial
manuscript above for the workflow. `cureAssess` is MIT licensed; authors Geethanjalee Mudunkotuwa
(author, creator, copyright holder) and Durbadal Ghosh (author); upstream
`https://github.com/GeethanjaleeM/cureAssess`. This app is a front end; every statistic is the
package&rsquo;s.

## Session information

```{r session-info, comment=NA}
sessionInfo()
```

```{r footer, results='asis'}
cat('<p class="ca-foot">Generated by cureAssessApp. Every inferential number above came out of a ',
    'field of a <span class="ca-mono">cureAssess</span> object; the app computed only the six ',
    'descriptive data counts and the <span class="ca-mono">qn</span> threshold ',
    '1 - α^(1/n), which <span class="ca-mono">qn.test()</span> does not return.</p>\n', sep = "")
```
