---
title: "Cure model appropriateness"
output:
  html_document:
    self_contained: true
    theme: null
    highlight: null
    mathjax: null
    toc: true
    toc_depth: 2
params:
  label: NULL
  source: NULL
  map: NULL
  prepared: NULL
  fit: NULL
  assess: NULL
  include_lognormal: FALSE
  expert: "unanswered"
---

```{r setup, include=FALSE}
# =============================================================================
# report/report.Rmd — the downloadable record of ONE assessment.
# Owner: builder-rec.
#
# THE CARDINAL RULE (S1) HOLDS HERE TOO. This document computes no inferential
# statistic. Every inferential number below is read out of a field of an object
# returned by an exported cureAssess function. The only arithmetic performed in
# this file is the app's own permitted list: the six descriptive data counts
# (n, events, censored %, median follow-up, max follow-up, last event time) and
# the single closed form 1 - alpha^(1/n) for the qn threshold, which qn.test()
# does not return as a field.
#
# SELF-CONTAINED, BY DESIGN. Nothing here is sourced from app/R/. The app
# renders this file from a copy in tempdir() with knit_root_dir set there, so
# no relative path reaches app/R/helpers.R, and an absolute path would break
# the promise that this file also knits by hand on any machine. The rp_*
# helpers therefore mirror helpers.R by hand, and the CSS is inlined rather
# than linked, so the output is one HTML file that opens offline.
#
# THE MIRROR THAT MUST NOT DRIFT: rp_recommendation() below is a hand copy of
# ca_recommendation() (V2_CONTRACT §C.3, app/R/helpers.R). Same four rules,
# same order, same headlines, same reason strings, same provisional flag. If
# one changes, both change, or the report and the screen start disagreeing —
# which is precisely the failure this pass was asked to close.
#
# V2_CONTRACT, this pass:
#   * §B.1 / §H.2.2 — the `assess` parameter is now the DEFAULT assessment,
#     handed over straight. The alpha, tau and distribution controls are
#     display-only and never reach this file, so the report always describes
#     the default alpha, the default tau and the default distribution. The
#     alpha printed below is still read off the package's own mz/shen objects,
#     never invented.
#   * THE THREE-WAY EXPERT STATE, RESOLVED. The old `expert_confirmed` boolean
#     could say only "confirmed" or "not confirmed", so this file could not
#     reproduce rule 1 and printed a numbers-only verdict where the app
#     printed "Do not use a cure model." The parameter is now `expert`, one of
#     "yes" / "no" / "unanswered", straight from ca_expert_state(state) — the
#     same value that reaches ca_recommendation() and the batch CSV. All four
#     rules are reproduced here and the screen and the report cannot disagree.
#
# FINAL_CONTRACT, this pass (owner builder-report):
#   * §P — the three follow-up cards now label BOTH a test statistic and a
#     p-value, read from the fields §P.2 names and from nowhere else. mz and qn
#     are algebraically one test (mz$statistic == (1 - qn$statistic)^n exactly),
#     so they share a p-value; one line under ## Diagnostics says so, once.
#     Shen reports no statistic and none is back-solved (§P.2.4).
#   * P2 — the Session information section is DELETED.
#   * P3 — rp_tech() and its five call sites are DELETED. The interpretation
#     strings go with them: each card already states its interpretation in the
#     report's own words, so the disclosure was duplication, and it was the last
#     place package field and model names reached the page.
#   * E2 / §G.2 — "whatever the numbers say" is struck from rule 1's reason.
#   * §F — this ONE template is also the PDF. mod_recommendation.R renders it to
#     HTML and prints that with headless Chrome, so there is no pdf_document
#     output format here and no second params block. Do not add either.
#
# S10 — the immune summary is gone from the output. The one surviving read is
# $tests$immune$last_observation_censored, in the three-checks chunk, with no
# immune branding, no statistic printed and no test framing. $tests_reason
# survives in the one place it is still needed: the diagnostics-were-not-run
# branch.
# =============================================================================

knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 9, fig.height = 6.5, dpi = 110
)

# ---- what did we actually get? ---------------------------------------------
# Every param has a NULL-safe default so this file knits by hand with no app
# running. Each section below is gated on one of these flags and degrades to a
# plain sentence rather than erroring.
P_PREPARED <- if (is.data.frame(params$prepared) && nrow(params$prepared) > 0L) params$prepared else NULL
P_ASSESS   <- params$assess
P_FIT      <- params$fit
P_MAP      <- params$map

HAVE_PREP   <- !is.null(P_PREPARED)
HAVE_ASSESS <- !is.null(P_ASSESS)
HAVE_TESTS  <- HAVE_ASSESS && isTRUE(P_ASSESS$tests_run) && !is.null(P_ASSESS$tests)
HAVE_AIC    <- HAVE_ASSESS && is.data.frame(P_ASSESS$screening$aic_table)
HAVE_KM     <- !is.null(P_FIT) && !is.null(P_FIT$kmplot)
HAVE_ANY    <- HAVE_PREP || HAVE_ASSESS

TESTS <- if (HAVE_TESTS) P_ASSESS$tests else NULL

# ---- formatting, identical in behaviour to helpers.R F.5-F.7 ---------------
rp_dash <- function() "—"

rp_num <- function(x, digits = 4) {
  if (is.null(x) || length(x) != 1L) return(rp_dash())
  v <- suppressWarnings(as.numeric(x))
  if (length(v) != 1L || !is.finite(v)) return(rp_dash())
  if (v != 0 && abs(v) < 1e-4) return(format(v, digits = 3, scientific = TRUE))
  formatC(v, format = "f", digits = digits)
}

rp_pct <- function(x, digits = 1) {
  if (is.null(x) || length(x) != 1L) return(rp_dash())
  v <- suppressWarnings(as.numeric(x))
  if (length(v) != 1L || !is.finite(v)) return(rp_dash())
  paste0(formatC(v, format = "f", digits = digits), "%")
}

# A length-1 finite numeric out of a package field, or NA. Package statistics
# come back NA when a diagnostic is not computable and NULL when tests_run is
# FALSE; both must land on an em dash, never on the text "NA".
rp_stat <- function(x) {
  if (is.null(x) || length(x) != 1L) return(NA_real_)
  v <- suppressWarnings(as.numeric(x))
  if (length(v) != 1L || !is.finite(v)) NA_real_ else v
}

esc <- function(x) htmltools::htmlEscape(as.character(x))

rp_chip <- function(variant, label) {
  stopifnot(variant %in% c("pass", "fail", "neutral", "caution", "void"))
  glyph <- c(pass = "✓", fail = "✕", neutral = "○",
             caution = "!", void = "—")[[variant]]
  paste0('<span class="ca-chip ca-chip--', variant, '"><span class="ca-chip__g">',
         glyph, '</span>', esc(label), '</span>')
}

# FINAL_CONTRACT P3 / §H: rp_tech() and every call to it are DELETED. The
# disclosures repeated an interpretation the card already states in its own
# words, and they were the last place package field and model names reached the
# page. The one string that survives is $tests_reason, printed as a plain <p> in
# the diagnostics-were-not-run branch and nowhere else.

rp_card <- function(title, body_html, chip = NULL, lede = NULL) {
  paste0(
    '<section class="ca-card"><header class="ca-card__head">',
    '<h3 class="ca-card__title">', esc(title), '</h3>',
    if (!is.null(chip)) chip else "",
    '</header>',
    if (!is.null(lede)) paste0('<p class="ca-card__lede">', esc(lede), '</p>') else "",
    '<div class="ca-card__body">', body_html, '</div>',
    '</section>'
  )
}

rp_stat_line <- function(value_txt, label = "statistic") {
  paste0('<p class="ca-stat"><span class="ca-stat__value">', esc(value_txt),
         '</span> <span class="ca-stat__label">', esc(label), '</span></p>')
}

rp_threshold_line <- function(txt) {
  paste0('<p class="ca-threshold"><span class="ca-threshold__value">', esc(txt),
         '</span></p>')
}

rp_kv_table <- function(rows) {
  cells <- vapply(rows, function(r) {
    paste0('<tr><th scope="row">', esc(r[[1]]), '</th><td>', esc(r[[2]]), '</td></tr>')
  }, character(1))
  paste0('<table class="ca-kv">', paste(cells, collapse = ""), '</table>')
}

# Display names for the ten model keys, mirroring helpers.R CA_MODEL_LABELS
# (REVISION_CONTRACT N.3) by hand, because helpers.R is not sourced here. An
# unknown key falls back to the key itself, never to NA and never to an error.
# The AIC table's own cells keep the package's raw keys: a table is data.
RP_MODEL_LABELS <- c(
  exponential       = "exponential",
  exponential_cure  = "exponential cure model",
  weibull           = "Weibull",
  weibull_cure      = "Weibull cure model",
  gamma             = "gamma",
  gamma_cure        = "gamma cure model",
  loglogistic       = "log-logistic",
  loglogistic_cure  = "log-logistic cure model",
  lognormal         = "lognormal",
  lognormal_cure    = "lognormal cure model"
)
rp_model_label <- function(key) {
  if (is.null(key) || length(key) != 1L || is.na(key)) return(NA_character_)
  k <- as.character(key)
  if (k %in% names(RP_MODEL_LABELS)) unname(RP_MODEL_LABELS[[k]]) else k
}

# The AIC table's parameter cell, relabelled for C1 exactly as
# mod_quantitative.R relabels it, so screen and report read the same. "theta"
# is the package's own name for the cure fraction (for the best-fit row its
# value is byte-identical to the pi-hat the RECeUS card reports); "shape",
# "scale" and "rate" are ordinary distribution vocabulary and are left alone.
# The leading "=" strip is the upstream one-parameter-fit bug the screen also
# works around. Display only: no number is altered.
rp_param_estimates <- function(x) {
  if (is.null(x) || length(x) != 1L || is.na(x)) return("\u2014")
  s <- sub("^=", "", as.character(x))
  gsub("(^|; )theta=", "\\1cure fraction=", s)
}

# ---- the six descriptive numbers (helpers.R F.1; permitted app-side) -------
rp_summary <- function(prepared) {
  if (is.null(prepared) || !is.data.frame(prepared) || nrow(prepared) == 0L ||
      is.null(prepared$Y) || is.null(prepared$D)) {
    return(list(n = 0L, events = 0L, censored_pct = NA_real_,
                median_followup = NA_real_, max_followup = NA_real_,
                last_event_time = NA_real_))
  }
  y <- as.numeric(prepared$Y); d <- as.integer(prepared$D)
  ev <- sum(d == 1L, na.rm = TRUE)
  list(
    n = length(y),
    events = ev,
    censored_pct = 100 * (1 - mean(d == 1L, na.rm = TRUE)),
    median_followup = stats::median(y, na.rm = TRUE),
    max_followup = max(y, na.rm = TRUE),
    last_event_time = if (ev > 0L) max(y[d == 1L], na.rm = TRUE) else NA_real_
  )
}

# ---- the one permitted closed form (helpers.R F.4) ------------------------
# qn.test() returns method, statistic and interpretation only: the threshold
# 1 - alpha^(1/n) sits inside the interpretation sentence and in no field, so
# the app (and this report) evaluate it. It is a closed form in alpha and the
# sample size alone. For qn, LARGER is better.
rp_qn_threshold <- function(n, alpha = 0.05) {
  n <- suppressWarnings(as.numeric(n)); a <- suppressWarnings(as.numeric(alpha))
  if (length(n) != 1L || !is.finite(n) || n < 1) return(NA_real_)
  if (length(a) != 1L || !is.finite(a) || a <= 0 || a >= 1) return(NA_real_)
  1 - a^(1 / n)
}

# The alpha in force — always the default one, since the exploration controls
# never reach this file (§B.1). Read off the package's own mz/shen objects,
# never invented.
# reads $tests$mz$alpha, then $tests$shen$alpha
ALPHA <- 0.05
if (HAVE_TESTS) {
  a <- rp_stat(TESTS$mz$alpha)
  if (is.na(a)) a <- rp_stat(TESTS$shen$alpha)
  if (!is.na(a)) ALPHA <- a
}
N_PREP <- if (HAVE_PREP) nrow(P_PREPARED) else NA_integer_

# ---- the three follow-up tests, as chips -----------------------------------
# DIRECTIONS, and getting qn's backwards is a blocking bug:
#   Maller-Zhou  statistic < alpha                  -> follow-up sufficient
#   qn           statistic > 1 - alpha^(1/n)        -> follow-up sufficient
#   Shen         statistic < alpha                  -> follow-up sufficient
# Maller-Zhou and qn are algebraically the same test on these data, so they
# can never disagree and are never independent confirmation of one another.
# reads $tests$mz$statistic, $tests$qn$statistic, $tests$shen$statistic and
# $tests$mz$alpha / $tests$shen$alpha
rp_followup <- function() {
  if (!HAVE_TESTS) return(NULL)
  a_mz <- rp_stat(TESTS$mz$alpha);   if (is.na(a_mz)) a_mz <- ALPHA
  a_sh <- rp_stat(TESTS$shen$alpha); if (is.na(a_sh)) a_sh <- ALPHA
  qn_thr <- rp_qn_threshold(N_PREP, ALPHA)

  mz <- rp_stat(TESTS$mz$statistic)
  qn <- rp_stat(TESTS$qn$statistic)
  sh <- rp_stat(TESTS$shen$statistic)

  list(
    list(name = "Maller-Zhou", stat = mz, threshold = a_mz, computable = !is.na(mz),
         sufficient = !is.na(mz) && mz < a_mz),
    list(name = "qn", stat = qn, threshold = qn_thr, computable = !is.na(qn),
         sufficient = !is.na(qn) && !is.na(qn_thr) && qn > qn_thr),
    list(name = "Shen", stat = sh, threshold = a_sh, computable = !is.na(sh),
         sufficient = !is.na(sh) && sh < a_sh)
  )
}
FU <- rp_followup()

# ---- the not-computable wording, one line (§G.5.2) -------------------------
RP_NOT_COMPUTABLE     <- "Cannot be computed: the longest observed time is an event."
RP_NOT_COMPUTABLE_REC <- "Cannot be computed: the model behind this number did not converge."

# FINAL_CONTRACT §P.2.8 — the card gained `pval`, defaulting to NA.
#
# The two labelled quantities and where each one comes from (§P.2, and these
# are the ONLY reads):
#
#   Maller-Zhou  p-value        $tests$mz$statistic    already on the p-value scale
#   Maller-Zhou  test statistic $tests$qn$statistic    a proportion
#   qn           test statistic $tests$qn$statistic
#   qn           p-value        $tests$mz$statistic
#   Shen         p-value        $tests$shen$statistic
#
# The cross-read is deliberate and is the whole point: mz$statistic is exactly
# (1 - qn$statistic)^n, Maller & Zhou (1994) eq. 5, so the two cards share one
# p-value by construction. This file NEVER computes (1 - qn)^n or 1 - mz^(1/n);
# it reads the field (§P.2.3). Shen's underlying proportion is not returned by
# the package, so no statistic is shown for Shen and none is back-solved
# (§P.2.4).
#
# `pval_first` follows S8: for Maller-Zhou and Shen smaller is better, so the
# p-value leads; for qn larger is better, so the statistic leads.
rp_followup_card <- function(title, stat, thr_text, sufficient,
                             pval = NA_real_, pval_first = FALSE,
                             show_stat = TRUE, note = NULL) {
  # S6 / §P.2.6: when the largest observed time is an event all three
  # diagnostics return NA together. No labelled rows are rendered at all —
  # never an em dash with a label beside it — and the literal text "NA" never
  # reaches the page.
  if (is.na(stat) && is.na(pval)) {
    return(rp_card(
      title,
      paste0('<p class="ca-void-line">', esc(RP_NOT_COMPUTABLE), '</p>'),
      chip = rp_chip("void", "Cannot be computed")
    ))
  }

  stat_row <- if (isTRUE(show_stat) && !is.na(stat)) {
    rp_stat_line(rp_num(stat), "test statistic")
  } else ""
  pval_row <- if (!is.na(pval)) rp_stat_line(rp_num(pval), "p-value") else ""

  body <- paste0(
    if (isTRUE(pval_first)) paste0(pval_row, stat_row) else paste0(stat_row, pval_row),
    rp_threshold_line(thr_text),
    if (!is.null(note)) paste0('<p class="ca-lede">', esc(note), '</p>') else ""
  )
  rp_card(
    title, body,
    chip = if (isTRUE(sufficient)) rp_chip("pass", "Follow-up sufficient")
           else rp_chip("fail", "Not sufficient")
  )
}

# ---- the expert judgment, three ways, from one parameter -------------------
# "yes" / "no" / "unanswered", straight from ca_expert_state(state). Anything
# unexpected — including a hand knit with no parameters at all — falls back to
# "unanswered", which is the honest reading of "nobody told this file".
EXPERT <- local({
  e <- params$expert
  if (is.null(e) || length(e) != 1L || is.na(e)) return("unanswered")
  e <- tolower(as.character(e))
  if (e %in% c("yes", "no", "unanswered")) e else "unanswered"
})

# ---- THE RECOMMENDATION — a hand mirror of ca_recommendation() -------------
# V2_CONTRACT §C.3. Four rules, in this order, and the order matters: an
# expert "no" settles the question whatever the numbers say.
#
# reads $screening$best_model_type and $tests$receus$decision — and NOTHING
# else (R8). The alpha slider, the tau slider, the distribution override and
# every individual follow-up statistic are display, not inputs; they do not
# reach this function and they do not reach this file.
#
# Returns list(headline, variant, reasons, provisional). `variant` is
# NA_character_ for "No recommendation.", which carries no banner colour: a
# verdict this report cannot name is a verdict it must not style either.
RP_DECISIONS <- c(
  "Cure model appropriate",
  "Cure model not supported",
  "Follow-up insufficient for cure modeling"
)

rp_recommendation <- function(aic, dec, expert = "unanswered") {

  # ---- Rule 1 — expert judgment overrides everything ----------------------
  if (identical(expert, "no")) {
    return(list(
      headline = "Do not use a cure model.",
      variant  = "ca-rec--unsupported",
      # E2 / §G.2 — the same string as ca_recommendation() rule 1 in
      # app/R/helpers.R and as the Expert judgment page's "no" outcome line.
      # All three, or the screen and the report disagree.
      reasons  = paste("You answered no at the expert judgment step. A cure model is not",
                       "appropriate here."),
      provisional = FALSE
    ))
  }

  # Rule 4 rides along with Rules 2 and 3, never with Rule 1. An expert "yes"
  # adds nothing at all: a silent pass is a pass.
  provisional <- identical(expert, "unanswered")

  # ---- Rule 2 — nothing was fitted ----------------------------------------
  if (is.null(aic) || length(aic) != 1L || is.na(aic)) {
    return(list(headline = "No recommendation.", variant = NA_character_,
                reasons = "No model could be fitted to these data.",
                provisional = provisional))
  }

  # An ABSENT or NA decision is "No recommendation.", not a crash: the
  # diagnostics can legitimately not run ($tests_run FALSE). The sentence is
  # ca_recommendation()'s own, character for character. An UNRECOGNISED
  # decision string still stops loudly, below.
  if (is.null(dec) || length(dec) != 1L || is.na(dec)) {
    return(list(headline = "No recommendation.", variant = NA_character_,
                reasons = "The diagnostics did not run for these data.",
                provisional = provisional))
  }
  dec <- as.character(dec)
  if (!(dec %in% RP_DECISIONS)) {
    stop("report.Rmd: unrecognised RECeUS decision string: ", paste(dec, collapse = ", "))
  }

  # ---- Rule 3 — the screening result crossed with the decision ------------
  cure <- identical(as.character(aic), "cure")
  also_non_cure <- "A model without a cured group also describes these data better."

  if (identical(dec, "Cure model appropriate")) {
    if (cure) {
      return(list(headline = "Use a cure model.", variant = "ca-rec--appropriate",
                  reasons = "There is a cured group, and follow-up is long enough to measure it.",
                  provisional = provisional))
    }
    return(list(headline = "Do not use a cure model.", variant = "ca-rec--unsupported",
                reasons = paste("Follow-up and the cured group look adequate, but a model without a",
                                "cured group describes these data better."),
                provisional = provisional))
  }

  if (identical(dec, "Follow-up insufficient for cure modeling")) {
    r <- paste("Follow-up is too short to tell a cured patient from one who has not",
               "had the event yet.")
  } else {
    r <- "The cured group is too small to be worth modelling."
  }
  if (!cure) r <- c(r, also_non_cure)
  list(headline = "Do not use a cure model.", variant = "ca-rec--unsupported",
       reasons = r, provisional = provisional)
}
```

<style type="text/css">
/* FIXPASS: inlined, never linked from app/www/app.css, so the output HTML is
   one self-contained file that opens offline. The tokens are the app's own.
   The step-strip, provenance and rec-condition rules went with the blocks
   §E.7 deleted. */
:root {
  --ca-surface:#FBFCFC; --ca-canvas:#EEF2F3; --ca-sunk:#F4F7F8;
  --ca-ink:#11191C; --ca-ink2:#414D52; --ca-ink3:#667378;
  --ca-rule:#DBE2E4; --ca-rule2:#C3CDD0; --ca-accent:#0A5670;
  --ca-pass-ink:#175A2B; --ca-pass-bg:#E6F3E9; --ca-pass-line:#B4D7BE;
  --ca-fail-ink:#8E2C20; --ca-fail-bg:#FBEAE6; --ca-fail-line:#EDC2B8;
  --ca-caution-ink:#7A4E05; --ca-caution-bg:#FBF1DE; --ca-caution-line:#E8D0A2;
  --ca-neutral-ink:#0A5670; --ca-neutral-bg:#E6F0F4; --ca-neutral-line:#BCD5E0;
  --ca-void-ink:#4B565A; --ca-void-bg:#ECEFF0; --ca-void-line:#CBD3D5;
  --ca-sans: system-ui, -apple-system, BlinkMacSystemFont, "Segoe UI", Roboto, "Helvetica Neue", Arial, sans-serif;
  --ca-serif: Charter, "Bitstream Charter", "Sitka Text", Cambria, "Iowan Old Style", "Palatino Linotype", Georgia, serif;
  --ca-mono: ui-monospace, SFMono-Regular, "SF Mono", Menlo, Consolas, "Liberation Mono", monospace;
}
body { font-family: var(--ca-sans); font-size: 16px; line-height: 1.55;
       color: var(--ca-ink); background: var(--ca-canvas);
       max-width: 62rem; margin: 0 auto; padding: 2rem 1.25rem 4rem; }
h1, h2, h3, h4 { font-family: var(--ca-serif); font-weight: 600; color: var(--ca-ink);
                 line-height: 1.25; }
h1 { font-size: 2rem; margin: 0 0 .25rem; }
h2 { font-size: 1.5rem; margin: 2.25rem 0 .75rem; padding-bottom: .4rem;
     border-bottom: 2px solid var(--ca-rule2); }
h3 { font-size: 1.2rem; margin: 0; }
h4 { font-size: 1.05rem; margin: 1.5rem 0 .5rem; }
a { color: var(--ca-accent); }
code, .ca-mono { font-family: var(--ca-mono); font-size: .9em; }
#TOC { background: var(--ca-surface); border: 1px solid var(--ca-rule);
       border-radius: 10px; padding: .75rem 1rem; margin: 1.5rem 0; }
#TOC ul { margin: .25rem 0; padding-left: 1.1rem; }

.ca-card { background: var(--ca-surface); border: 1px solid var(--ca-rule);
           border-radius: 10px; padding: 1rem 1.15rem; margin: 0 0 1rem; }
.ca-card__head { display: flex; flex-wrap: wrap; align-items: baseline;
                 justify-content: space-between; gap: .5rem; margin-bottom: .5rem; }
.ca-card__lede { color: var(--ca-ink2); margin: 0 0 .6rem; }
.ca-card__body > :first-child { margin-top: 0; }
.ca-card__body > :last-child { margin-bottom: 0; }

.ca-chip { display: inline-block; border: 1px solid; border-radius: 999px;
           padding: .15rem .65rem; font-size: .9rem; font-weight: 600; }
.ca-chip__g { margin-right: .4rem; font-weight: 700; }
.ca-chip--pass    { color: var(--ca-pass-ink);    background: var(--ca-pass-bg);    border-color: var(--ca-pass-line); }
.ca-chip--fail    { color: var(--ca-fail-ink);    background: var(--ca-fail-bg);    border-color: var(--ca-fail-line); }
.ca-chip--caution { color: var(--ca-caution-ink); background: var(--ca-caution-bg); border-color: var(--ca-caution-line); }
.ca-chip--neutral { color: var(--ca-neutral-ink); background: var(--ca-neutral-bg); border-color: var(--ca-neutral-line); }
.ca-chip--void    { color: var(--ca-void-ink);    background: var(--ca-void-bg);    border-color: var(--ca-void-line); font-weight: 500; }

.ca-stat { margin: .3rem 0 .4rem; }
.ca-stat__value { font-size: 1.7rem; font-weight: 700; font-family: var(--ca-mono); }
.ca-stat__label { color: var(--ca-ink3); font-size: .85rem; text-transform: uppercase;
                  letter-spacing: .04em; }
.ca-threshold { margin: 0 0 .5rem; color: var(--ca-ink2); }
.ca-threshold__value { font-family: var(--ca-mono); }
.ca-void-line { color: var(--ca-void-ink); }

.ca-tech { margin: .5rem 0 0; border: 1px solid var(--ca-rule); border-radius: 6px;
           background: var(--ca-sunk); }
.ca-tech > summary { cursor: pointer; padding: .4rem .7rem; font-size: .9rem;
                     color: var(--ca-ink2); }
.ca-tech__body { padding: 0 .7rem .6rem; }
.ca-tech__body p { margin: 0; }

.ca-statrow { display: flex; flex-wrap: wrap; gap: .75rem; }
.ca-statrow__cell { flex: 1 1 8.5rem; background: var(--ca-surface);
                    border: 1px solid var(--ca-rule); border-radius: 8px; padding: .6rem .75rem; }
.ca-statrow__label { display: block; font-size: .75rem; text-transform: uppercase;
                     letter-spacing: .05em; color: var(--ca-ink3); }
.ca-statrow__value { display: block; font-size: 1.45rem; font-weight: 700;
                     font-family: var(--ca-mono); }
.ca-statrow__note { display: block; font-size: .78rem; color: var(--ca-ink3); }

table.ca-kv { border-collapse: collapse; width: 100%; }
table.ca-kv th, table.ca-kv td { border-bottom: 1px solid var(--ca-rule);
                                 padding: .35rem .5rem; text-align: left; vertical-align: top; }
table.ca-kv th { width: 34%; font-weight: 600; color: var(--ca-ink2); }

.ca-tablewrap { overflow-x: auto; }
table.ca-aic { border-collapse: collapse; width: 100%; font-size: .9rem;
               background: var(--ca-surface); }
table.ca-aic th, table.ca-aic td { border: 1px solid var(--ca-rule); padding: .35rem .5rem;
                                   text-align: left; vertical-align: top; }
table.ca-aic thead th { background: var(--ca-sunk); }
table.ca-aic td.num { font-family: var(--ca-mono); text-align: right; white-space: nowrap; }
table.ca-aic td.err { color: var(--ca-fail-ink); }
.ca-badge { display: inline-block; border: 1px solid var(--ca-rule2); border-radius: 4px;
            padding: 0 .35rem; font-size: .8rem; }

.ca-note { border: 1px solid var(--ca-rule); border-left-width: 5px; border-radius: 8px;
           background: var(--ca-surface); padding: .8rem 1rem; margin: 1rem 0; }
.ca-note__title { font-family: var(--ca-serif); font-weight: 600; margin: 0 0 .35rem; }
.ca-note--info { border-left-color: var(--ca-neutral-ink); }
.ca-note--warning { border-left-color: var(--ca-caution-ink); background: var(--ca-caution-bg); }

.ca-checks p { margin: .35rem 0; }
.ca-rec { border: 1px solid var(--ca-rec-line, var(--ca-rule)); border-radius: 10px;
          background: var(--ca-rec-bg, var(--ca-surface)); padding: 1rem 1.15rem; margin: 1rem 0; }
.ca-rec--appropriate { --ca-rec-line: var(--ca-pass-line); --ca-rec-bg: var(--ca-pass-bg); --ca-rec-ink: var(--ca-pass-ink); }
.ca-rec--unsupported { --ca-rec-line: var(--ca-fail-line); --ca-rec-bg: var(--ca-fail-bg); --ca-rec-ink: var(--ca-fail-ink); }
.ca-rec__decision { font-family: var(--ca-serif); font-size: 1.65rem; font-weight: 700;
                    color: var(--ca-rec-ink, var(--ca-ink)); margin-bottom: .6rem; }
.ca-rec__body p { margin: .35rem 0; }
.ca-lede { color: var(--ca-ink2); }
pre { background: var(--ca-sunk); border: 1px solid var(--ca-rule); border-radius: 6px;
      padding: .75rem; overflow-x: auto; font-size: .82rem; }
@media print { body { background: #fff; } .ca-tech { display: none; } }
</style>

```{r title-block, results='asis'}
lbl <- if (!is.null(params$label) && nzchar(as.character(params$label)[1])) {
  as.character(params$label)[1]
} else "no dataset supplied"
cat('<p class="ca-lede">Dataset: <strong>', esc(lbl), '</strong> · generated ',
    esc(format(Sys.time(), "%Y-%m-%d %H:%M")), ' by cureAssessApp.</p>\n', sep = "")
cat('<p>cureAssessApp decides whether a cure model is appropriate for right-censored ',
    'survival data. It does not fit your final model and it does not estimate treatment ',
    'effects.</p>\n', sep = "")
```

```{r nothing-supplied, results='asis', eval=!HAVE_ANY}
cat('<div class="ca-note ca-note--warning"><p class="ca-note__title">No assessment supplied</p>',
    '<p>This report was knitted with its default parameters, so there is no dataset and no ',
    'assessment to describe.</p></div>\n', sep = "")
```

## Dataset

```{r dataset, results='asis'}
# All six rows are data-handling facts held in state$map, not statistics.
src <- params$source
src_txt <- if (identical(src, "builtin")) {
  "Built-in example shipped with the app"
} else if (identical(src, "upload")) {
  "Uploaded by the user"
} else {
  rp_dash()
}
scale_txt <- if (!is.null(P_MAP$time_scale)) {
  if (identical(P_MAP$time_scale, "days_to_years")) {
    "Days converted to years: times were divided by 365.25"
  } else {
    "None: times were used as supplied"
  }
} else rp_dash()

rows <- list(
  c("Dataset", lbl),
  c("Where it came from", src_txt),
  c("Time column", if (!is.null(P_MAP$time)) P_MAP$time else rp_dash()),
  c("Status column", if (!is.null(P_MAP$status)) P_MAP$status else rp_dash()),
  c("Value meaning “event”", if (!is.null(P_MAP$event_level)) P_MAP$event_level else rp_dash()),
  c("Time scale", scale_txt)
)
cat(rp_card(
  "How this dataset was read",
  paste0(
    rp_kv_table(rows),
    '<p class="ca-lede">Every count here is after dropping rows with a missing time or status.</p>'
  )
), "\n")
```

## Data summary

```{r data-summary, results='asis'}
if (!HAVE_PREP) {
  cat('<p class="ca-lede">No prepared dataset was supplied, so there is nothing to summarise.</p>\n')
} else {
  # The six descriptive counts, computed here and permitted to be (F.1).
  s <- rp_summary(P_PREPARED)
  cells <- list(
    c("n", format(s$n, big.mark = ","), "rows analysed"),
    c("Events", format(s$events, big.mark = ","), "event observed"),
    c("Censored", rp_pct(s$censored_pct), "share of rows"),
    c("Median follow-up", rp_num(s$median_followup, 3), "median observed time"),
    c("Max follow-up", rp_num(s$max_followup, 3), "largest observed time"),
    c("Last event", rp_num(s$last_event_time, 3), "largest event time")
  )
  html <- paste0(vapply(cells, function(c3) paste0(
    '<div class="ca-statrow__cell"><span class="ca-statrow__label">', esc(c3[1]),
    '</span><span class="ca-statrow__value">', esc(c3[2]),
    '</span><span class="ca-statrow__note">', esc(c3[3]), '</span></div>'
  ), character(1)), collapse = "")
  cat('<div class="ca-statrow">', html, '</div>\n', sep = "")
}
```

## Kaplan-Meier curve

```{r km-lede, results='asis', eval=HAVE_KM}
cat('<p class="ca-lede">The shaded band runs from the last event to the end of follow-up, and ',
    'the table underneath is the number of subjects still under observation.</p>\n', sep = "")
```

```{r km-draw, fig.alt="Kaplan-Meier survival curve with a number-at-risk table underneath", eval=HAVE_KM}
# reads $kmplot from model.fitting(plot_km = TRUE)
ok <- tryCatch({
  if (requireNamespace("survminer", quietly = TRUE)) invisible(NULL)
  print(P_FIT$kmplot)
  TRUE
}, error = function(e) {
  message("KM plot could not be drawn: ", conditionMessage(e))
  FALSE
})
```

```{r km-missing, results='asis', eval=!HAVE_KM}
cat('<p class="ca-lede">No Kaplan-Meier plot was supplied, so there is no curve to draw.</p>\n')
```

## Model screening

```{r aic-table, results='asis'}
if (!HAVE_AIC) {
  cat('<p class="ca-lede">No assessment was supplied, so there is no model comparison.</p>\n')
} else {
  # reads $screening$aic_table. The error column is always shown and failed
  # rows are never removed: a failed fit sets AIC = NA and records its reason
  # there, so hiding the column would turn a silent failure into an invisible
  # one.
  tb <- P_ASSESS$screening$aic_table
  n_models <- nrow(tb)
  set_txt <- if (isTRUE(params$include_lognormal)) {
    paste0(n_models, " candidate models, with the lognormal pair included.")
  } else {
    paste0(n_models, " candidate models, without the lognormal pair.")
  }
  cat('<p class="ca-lede">', esc(set_txt), '</p>\n', sep = "")

  hdr <- paste0('<tr><th>Model</th><th>Type</th><th>AIC</th>',
                '<th>Parameters</th><th>Problem</th></tr>')
  body <- vapply(seq_len(n_models), function(i) {
    paste0(
      # C1 / screen-report agreement: the MODEL cell carries the display name,
      # exactly as mod_quantitative.R renders it, not the package key
      # ("log-logistic cure model", not "loglogistic_cure"). A relabelling
      # only: no row is added, removed or reordered, and no number moves.
      '<tr><td><span class="ca-mono">', esc(rp_model_label(tb$model[i])), '</span></td>',
      '<td><span class="ca-badge">', esc(tb$model_type[i]), '</span></td>',
      '<td class="num">', esc(rp_num(tb$AIC[i], 2)), '</td>',
      # Same C1 relabel of the "theta" KEY the screen applies: it is the
      # package's internal name for the cure fraction. Key only, anchored, so
      # no number and no other parameter name is touched.
      '<td><span class="ca-mono">', esc(rp_param_estimates(tb$parameter_estimates[i])), '</span></td>',
      '<td class="err">', esc(tb$error[i]), '</td></tr>'
    )
  }, character(1))
  cat('<div class="ca-tablewrap"><table class="ca-aic"><thead>', hdr, '</thead><tbody>',
      paste(body, collapse = ""), '</tbody></table></div>\n', sep = "")

  # reads $screening$best_model, $screening$best_model_type and
  # $screening$initial_decision
  best <- P_ASSESS$screening$best_model
  btyp <- P_ASSESS$screening$best_model_type
  best_lbl <- rp_model_label(best)
  cat(rp_card(
    "Best fit",
    paste0(
      if (is.na(best_lbl)) {
        '<p>No model could be fitted to these data.</p>'
      } else {
        paste0('<p><strong>Best fit: ', esc(best_lbl), '.</strong> ',
               '<span class="ca-badge">',
               esc(if (length(btyp) == 1L && !is.na(btyp)) btyp else rp_dash()),
               '</span></p>')
      }
    )
  ), "\n")
}
```

## Diagnostics

```{r diag-cards, results='asis', eval=HAVE_TESTS}
cat('<p class="ca-lede">Is there strong quantitative evidence of sufficient follow-up and a ',
    'cure fraction?</p>\n', sep = "")

# §P.2.5 — one short line, once, because two cards now visibly carry the same
# p-value. It is not repeated on either card.
cat('<p class="ca-lede">Maller-Zhou and qn are two readings of one test, so they ',
    'share a p-value and cannot disagree.</p>\n', sep = "")

qn_thr <- rp_qn_threshold(N_PREP, ALPHA)

MZ_P <- rp_stat(TESTS$mz$statistic)     # the shared p-value
QN_S <- rp_stat(TESTS$qn$statistic)     # the shared test statistic

# Maller-Zhou: SMALLER is better (S8), so the p-value leads.
cat(rp_followup_card(
  "Maller-Zhou test",
  stat = QN_S, pval = MZ_P, pval_first = TRUE,
  thr_text   = paste0("Sufficient if the p-value is below ", rp_num(FU[[1]]$threshold, 3)),
  sufficient = FU[[1]]$sufficient
), "\n")

# qn: LARGER is better (S8), against the app-computed threshold 1 - alpha^(1/n),
# so the statistic leads.
cat(rp_followup_card(
  "qn statistic",
  stat = QN_S, pval = MZ_P, pval_first = FALSE,
  thr_text   = paste0("Sufficient if the statistic is above ", rp_num(qn_thr, 7)),
  sufficient = FU[[2]]$sufficient
), "\n")

# Shen: the package returns method, statistic, alpha and interpretation, and
# nothing else. The underlying count and proportion are NOT returned, so no
# statistic is shown and none is recovered (§P.2.4).
cat(rp_followup_card(
  "Shen test",
  stat = NA_real_, show_stat = FALSE,
  pval = rp_stat(TESTS$shen$statistic), pval_first = TRUE,
  thr_text   = paste0("Sufficient if the p-value is below ", rp_num(FU[[3]]$threshold, 3)),
  sufficient = FU[[3]]$sufficient,
  note       = "Only the p-value is reported for this test."
), "\n")

# RECeUS. reads $tests$receus$pi_hat, $r_hat, $tau, $decision, $interpretation
rc <- TESTS$receus
pi_hat <- rp_stat(rc$pi_hat); r_hat <- rp_stat(rc$r_hat); tau <- rp_stat(rc$tau)
if (is.na(pi_hat) && is.na(r_hat)) {
  cat(rp_card(
    "RECeUS",
    paste0(
      rp_stat_line(rp_dash(), "cure fraction"), rp_stat_line(rp_dash(), "remaining-uncured ratio"),
      '<p class="ca-void-line">', esc(RP_NOT_COMPUTABLE_REC), '</p>'
    ),
    chip = rp_chip("void", "Cannot be computed")
  ), "\n")
} else {
  # The chip label is the package's decision string, verbatim, with the fixed
  # variant lookup. An unrecognised string must not be styled as a verdict.
  chip_variant <- if (identical(rc$decision, "Cure model appropriate")) "pass" else "fail"
  cat(rp_card(
    "RECeUS",
    paste0(
      rp_stat_line(rp_num(pi_hat), "cure fraction"),
      rp_stat_line(rp_num(r_hat), "remaining-uncured ratio"),
      rp_threshold_line("Above 0.025 and below 0.05"),
      '<p class="ca-lede">Evaluated at ', esc(rp_num(tau)), ', the largest observed time.</p>'
    ),
    chip = rp_chip(chip_variant, rc$decision)
  ), "\n")
}
```

```{r diag-not-run, results='asis', eval=(HAVE_ASSESS && !HAVE_TESTS)}
# reads $tests_reason
# P3: the one surviving package string, printed plainly with no disclosure
# around it.
cat('<h3>Diagnostics were not run</h3>\n')
cat('<p>', esc(P_ASSESS$tests_reason), '</p>\n', sep = "")
```

```{r diag-none, results='asis', eval=!HAVE_ASSESS}
cat('<p class="ca-lede">No assessment was supplied, so there are no diagnostics.</p>\n')
```

## The three checks

```{r three-checks, results='asis'}
# Line 1 — expert judgment, all three states. This is ca_expert_state(state)
# arriving intact, so this line and the Expert judgment tab's own chip say the
# same thing about the same session.
l1 <- switch(EXPERT,
             yes = "answered yes",
             no  = "answered no",
             "not answered")

# Line 2 — visual assessment.
# reads $tests$immune$last_observation_censored; see REVISION_CONTRACT §F. This
# is a property of the data, not an inferential statistic, and nothing about
# where it came from is printed.
l2 <- if (!HAVE_TESTS) {
  NA_character_
} else if (isTRUE(TESTS$immune$last_observation_censored)) {
  "the last observed time is censored, so a plateau is possible"
} else {
  "the last observed time is an event, so there is no clear plateau"
}

# Line 3 — quantitative assessment: the decision string, verbatim.
# reads $tests$receus$decision
l3 <- if (HAVE_TESTS && !is.null(TESTS$receus$decision)) TESTS$receus$decision else NA_character_

lines <- c(
  paste0("<strong>Expert judgment:</strong> ", esc(l1), "."),
  if (!is.na(l2)) paste0("<strong>Visual assessment:</strong> ", esc(l2), ".") else
    "<strong>Visual assessment:</strong> not available.",
  if (!is.na(l3)) paste0("<strong>Quantitative assessment:</strong> ", esc(l3), ".") else
    "<strong>Quantitative assessment:</strong> not available."
)
cat('<div class="ca-checks">', paste0("<p>", lines, "</p>", collapse = ""), '</div>\n', sep = "")
```

## Recommendation

```{r recommendation, results='asis', eval=HAVE_ASSESS}
# The same four rules, the same headlines and the same reason wording as the
# app, read from the same two package fields plus the same expert answer.
#
# Rendered whenever an assessment was supplied, not only when the diagnostics
# ran: a run that produced no decision is a verdict of "No recommendation."
# (§C.3), which is what the screen shows too. An empty page here where the
# screen shows a banner would be exactly the disagreement this pass closes.
rec <- rp_recommendation(
  P_ASSESS$screening$best_model_type,
  if (HAVE_TESTS) TESTS$receus$decision else NULL,
  EXPERT
)

body <- paste0("<p>", vapply(rec$reasons, esc, character(1)), "</p>", collapse = "")
if (isTRUE(rec$provisional)) {
  body <- paste0(body,
                 '<p>Expert judgment is not answered yet, so this is provisional.</p>')
}
cat('<div class="ca-rec ', if (is.na(rec$variant)) "" else rec$variant, '">',
    '<div class="ca-rec__decision">', esc(rec$headline), '</div>',
    '<div class="ca-rec__body">', body, '</div>',
    '</div>\n', sep = "")
```

```{r recommendation-none, results='asis', eval=!HAVE_ASSESS}
cat('<p class="ca-lede">No assessment was supplied, so there is no recommendation.</p>\n')
```

## References

- Maller RA, Zhou S (1992). Estimating the proportion of immunes in a censored sample. *Biometrika*, 79(4), 731&ndash;739. doi:10.1093/biomet/79.4.731
- Maller RA, Zhou S (1994). Testing for sufficient follow-up and outliers in survival data. *Journal of the American Statistical Association*, 89(428), 1499&ndash;1506. doi:10.1080/01621459.1994.10476889
- Maller RA, Zhou S (1995). Testing for the presence of immune or cured individuals. *Biometrics*, 51, 1197&ndash;1205. doi:10.2307/2533253
- Maller RA, Zhou X (1996). *Survival Analysis with Long-Term Survivors*. Wiley.
- Shen P-S (2000). Testing for sufficient follow-up in survival data. *Statistics & Probability Letters*, 49(4), 313&ndash;322. doi:10.1016/S0167-7152(00)00063-8
- Escobar-Bach M, Van Keilegom I (2019). Non-parametric cure rate estimation under insufficient follow-up by using extremes. *Journal of the Royal Statistical Society Series B*, 81(5), 861&ndash;880.
- Othus M, Bansal A, Koepl L, Wagner S, Ramsey S (2020). Bias in mean survival from fitting cure models with limited follow-up. *Value in Health*, 23(8), 1034&ndash;1039.
- Selukar S, Othus M (2023). RECeUS: Ratio estimation of censored uncured subjects, a different approach for assessing cure model appropriateness in studies with long-term survivors. *Statistics in Medicine*, 42(3), 209&ndash;227. doi:10.1002/sim.9610
- Maller RA, Resnick S, Shemehsavar S (2024). Finite sample and asymptotic distributions of a statistic for sufficient follow-up in cure models. *Canadian Journal of Statistics*, 52(2), 359&ndash;379. doi:10.1002/cjs.11771
- Yuen TP, Musta E (2024). Testing for sufficient follow-up in survival data with a cure fraction. arXiv:2403.16832.
- Mudunkotuwa G, Ghosh D, Selukar S (2026). A Tutorial for Evaluating Cure Model Appropriateness. arXiv:2605.04999v2 [stat.ME].

Every statistic above is a field of a `cureAssess` object. `cureAssess` is MIT licensed;
authors Geethanjalee Mudunkotuwa and Durbadal Ghosh. Cite the tutorial manuscript for the
workflow this report follows.
