---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# cureAssess

<!-- badges: start -->
<!-- badges: end -->

cureAssess helps you decide whether a **cure model** is appropriate for
right-censored survival data — that is, whether the data plausibly contain a
fraction of subjects who will never experience the event of interest.

Fitting a cure model when follow-up is too short, or when there is no real
cured fraction, produces biased estimates. cureAssess packages the standard
screening and diagnostic procedures for that decision into one workflow.

## Installation

Install the released version from CRAN:

``` r
install.packages("cureAssess")
```

Or the development version from GitHub:

``` r
# install.packages("pak")
pak::pak("GeethanjaleeM/cureAssess")
```

## The two-stage workflow

**Stage 1 — screening.** Standardize the data, fit matched cure and non-cure
parametric models, and rank them by AIC. If the smallest-AIC model is a cure
model, that is initial support for cure modeling.

**Stage 2 — diagnostics.** Run the formal checks for sufficient follow-up and
for a cured fraction: the Maller-Zhou statistics, Shen's test, an immune
summary of the Kaplan-Meier tail, and the RECeUS method.

`cure.appropriateness()` runs both stages in one call.

## Example

```{r example}
library(cureAssess)
library(survival)

res <- cure.appropriateness(
  data = gbsg,
  time = "rfstime",
  status = "status",
  time_scale = "days_to_years",
  plot_km = FALSE,
  run_tests = "yes"
)

res
```

The AIC comparison behind the decision is available directly:

```{r aic-table}
head(res$screening$aic_table[, c("model", "model_type", "AIC")], 4)
```

Individual diagnostics can also be run on their own, on data prepared by
`prepare.surv.data()`:

```{r diagnostics}
dat <- prepare.surv.data(
  data = gbsg,
  time = "rfstime",
  status = "status",
  time_scale = "days_to_years"
)

mz.test(dat)
receus.method(dat, dist = "lnorm")
```

Note what the two stages say here. AIC preferred a *cure* model for these
data, but RECeUS reports that too large a share of uncured subjects is still
censored at the end of follow-up. A better AIC fit is not on its own evidence
that a cure fraction can be estimated reliably — which is why the screening
stage alone is not enough.

For a worked comparison of a data set where a cure model *is* supported
(`nwtco`) against one where it is *not* (`gbsg`), see the introductory
vignette:

``` r
vignette("introduction", package = "cureAssess")
```

## Interpreting the output

These diagnostics are descriptive aids, not a single decision rule. Sufficient
follow-up tests ask whether the study ran long enough to tell a genuine cure
fraction apart from a plateau caused by censoring. Read them together, and
alongside subject-matter knowledge about whether cure is clinically plausible.

## References

- Maller RA, Zhou S (1992). Estimating the proportion of immunes in a censored
  sample. *Biometrika*, 79(4), 731–739.
  <doi:10.1093/biomet/79.4.731>
- Maller RA, Zhou S (1994). Testing for sufficient follow-up and outliers in
  survival data. *Journal of the American Statistical Association*, 89(428),
  1499–1506. <doi:10.1080/01621459.1994.10476889>
- Shen P-S (2000). Testing for sufficient follow-up in survival data.
  *Statistics & Probability Letters*, 49(4), 313–322.
  <doi:10.1016/S0167-7152(00)00063-8>
- Selukar S, Othus M (2023). RECeUS: Ratio estimation of censored uncured
  subjects. *Statistics in Medicine*, 42(3), 209–227.
  <doi:10.1002/sim.9610>

## License

MIT © Geethanjalee Mudunkotuwa
